\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-01-12 honors_thesis/gecco_paper/paper.tex
Added another branch of this stuff in the form of a GECCO paper - I should have found a better way to do this...

\documentclass[conference,letterpaper]{IEEEtran}

@@

\IEEEoverridecommandlockouts

@@

We also propose another way of making use of these speculative particles,
keeping the best position found instead of the position that PSO actually would
have taken.  We show results that for a number of functions, speculative
evaluation can have dramatic improvements over adding additional particles to
the swarm.

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05},
and formulating PSO in Google's MapReduce framework \cite{mcnabb-cec07}.
Parallelizing the evaluation of the objective function can also be done in some
cases, though that is not an adaption of the PSO algorithm and thus is not the
focus of this paper.

@@

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait, guess which way the branch will go and start
executing.  If the processor guessed right, execution is much farther ahead
than if it had just idly waited on the branch.  If it guessed wrong, execution
restarts right where it would have been anyway.  This is called speculative
execution or branch prediction.

@@

Because the motion of particles in PSO depends only on the \emph{positions} of
the particle and its neighbors, not the \emph{value} of the objective function
at those positions, speculative evaluation is possible.  We do not have to wait
for a particle's position to be evaluated to know where it might move to next,
so we can compute the set of all possible next positions and send them to be
evaluated along with the particle's current position.  Once the current
position of each particle has been evaluated, we can determine which of the
speculatively evaluated positions matches the branch taken by the particle.  We
have thus evaluated two iterations at once.  This can easily get unwieldy if
the set of next possible positions for each particle is large, but a wise
choice of topology limits that set to a reasonable size.

The original PSO can be reproduced exactly, two iterations at a time, using
this technique.  However, in strictly using speculative evaluation to reproduce
the original PSO, a lot of positions are speculatively evaluated whose values
get ignored after deciding which branch was the correct one.  A better use of
those speculative evaluations may be to incorporate the information into the
swarm, no longer exactly reproducing PSO but perhaps producing better results.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
a multi-dimensional space by mimicking the swarming and flocking behavior of
birds and other animals. It is a social algorithm that depends on interaction
between particles to quickly and consistently find the optimal solution to some
objective function.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Table~\ref{tab:evals} lists all possible combinations of updates
that can be seen by a particle with two neighbors other than itself, which is
the case in all of the topologies we use.  If all of these speculative
positions are evaluated in parallel along with the original, non-speculative
position of the particle at time $t-1$, then one of the speculative positions
has to be the correct one, and two iterations have been evaluated at once.

\begin{table}
  \caption{All possible updates for a particle with two neighbors}
  \label{tab:evals}
  \centering
  \begin{tabular}{c|c}
	$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	No update&No update\\
	Updated&No update\\
	No update&Neighbor 1\\
	No update&Neighbor 2\\
	Updated&The particle itself\\
	Updated&Neigbhor 1\\
	Updated&Neighbor 2\\
  \end{tabular}
\end{table}

Because $2(n+1)$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require on
the order of $p^2$ speculative evaluations per iteration, where $p$ is the
swarm size.  Clearly it is much more desirable to have a sparse topology, where
$O(np)$ is much smaller than $O(p^2)$.  However, some functions are better
optimized with the Complete topology and the quick spread of information it
entails than with sparse topologies.  Accordingly, we use the Random topology
described in~\cite{mcnabb-cec09}, which has been shown to effectively simulate
the Complete topology.  In that topology, 2 neighbors are chosen at random from
the swarm for each particle at each iteration.  Because a particle is also
always its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report
the results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

@@

To aid in describing our methodology, we introduce a few terms.  A particle's
set of \emph{speculative children} is the set of all possible next iteration
states that a particle could have.  We use $\p_t$ to denote a particle at
iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
corresponding to one of the rows in Table~\ref{tab:evals}.  $\n_t$ is a
neighbor of particle $\p_t$.  A set of particles is $\pset$, $\sset$, or
$\nset$, whereas a single particle is simply $\p$, $\s$, or $\n$.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
a scheme, the master generates the positions to evaluate speculatively and
decides which one to accept for each particle.

@@

This is done by first updating each particle $\noeval{\p}_t$ to
$\nonbest{\p}_t$ and each speculative particle $\noeval{\s}_{t+1}$ to
$\nonbest{\s}_{t+1}$ by associating the value determined by the slave to the
particle it came from, and updating that particle's personal best if the value
reported is better than its previous personal best value.  Then each
$\nonbest{\p}_t$ must be updated to $\p_t$ by updating its neighborhood best
position and value with information about each of its neighbors.  This is
simply the original PSO algorithm.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depend entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected.

@@

The parent also needs to pass the value of the neighborhood best that the child
guessed.  The child only knows the position and needs the value in order to
make future comparisons between neighborhood best positions.

Upon picking the correct branch for each particle and updating the children's
personal best and neighborhood best value (from iteration $t$), the result is
the set $\nonbest{\pset}_{t+1}$, as the particles are now no longer
speculative.  What remains is to update the neighborhood best of those
particles from their neighbors (from iteration $t+1$), as above, to obtain
$\pset_{t+1}$.  That set of particles can subsequently be used to produce the
sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$, and the process repeats
itself.

The algorithm is outlined in \alg{centralized}.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those particles.
Messages such as values and positions for the neighborhood best are sent
between processors.  There may be some machine that collects information from
all of the particles and outputs the result of the algorithm, though that
machine's importance is much less than in centralized algorithms.

To perform speculative evaluation in a distributed PSO algorithm, there must be
some way to have processors evaluate the speculative children of particles,
without giving the speculative particles the same treatment as actual
particles, as the speculative children only live for one iteration.  Perhaps
that can be done by assigning each particle a set of machines instead of a
single machine, and the particle directs its extra machines to evaluate its
speculative children.  No matter the distributed framework used the message
passing should be similar, and that is what we describe here.

@@

In some distributed frameworks, like the MapReduce framework that we used in
our experiments, rounds of message passing can be expensive.  The method
just described uses three rounds of message passing for every two iterations.
Accordingly, we describe a method that requires only one round of communication
for each pair of iterations.  A lot more messages are needed, but that is
sometimes more desirable than synchronizing all of the machines three times.

@@

With $\nonbest{\nset}_{t+1}$ and $\nonbest{\p}_{t+1}$, we can produce
$\p_{t+1}$.  This is shown in \fig{messages4}.  Note that we just combined
Figures~\ref{fig:messages2}~and~\ref{fig:messages3}, putting them together
to make $\p_{t+1}$, as all the particle needs is it's neighborhood best to be
updated.

@@

This may seem like an inordinate amount of work, and in some instances it is.
However, some parallel frameworks necessitate this type of message passing, so
we have described how speculative evaluation can be performed in such
frameworks.

@@

Performing speculative evaluation in PSO with a dynamic topology raises a few
issues of its own.  First, with neighbors changing each iteration, messages
that processors pass to their neighbors need to be sent to the correct
neighbors for each iteration.  In centralized algorithms and in distributed
algorithms with several rounds of message passing (the first method described
above), nothing needs to be changed.  However, in the case where processors
reconstruct information about neighboring particles from messages, those
messages need to be sent to the correct neighbors.  A particle cannot simply
send messages to its neighbors' neighbors' neighbors---it needs to send
messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
For every neighbor outward information is sent, the iteration also needs to be
incremented, as information about neighbors' neighbors is used during iteration
$t+1$, and information about neighbors' neighbors' neighbors is used to
reconstruct information about iteration $t+2$.  Also, this method of message
passing again requires the use of random seeds if the topology is random.

@@

In a dynamic topology, a particle might not have information about the previous
positions of its neighbors at iteration $t$.  That means that its new
neighborhood best could come not only from its neighbors' positions at
iteration $t$, but also from their personal best from iteration $t-1$, as
neighbors' personal bests are what are used to update a particle's neighborhood
best.  That creates a problem for speculative evaluation---there are
potentially more than $2(n+1)$ possible next positions, increasing the amount
of work that must be done to perform the second iteration at the same time as
the first.

This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
available information about $\noeval{\nset}_{t+2}$ before producing
$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its 
neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
still only $2(n+1)$ possible next positions, and the problem is avoided.

@@

In performing speculative evaluation as we have described it, $2(n+1)$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  It seems reasonable to try to make use of those
evaluations instead of throwing them away.  Making use of them changes the
behavior of PSO, instead of reproducing it exactly as the above method
explains, but the change could be an improvement.

@@

In our experiments we compared our speculative PSO algorithm to the standard
PSO algorithm.  To use the same number of processors for each algorithm,
speculative evaluation needs a smaller swarm size than standard PSO.  For each
of the topologies we used, a particle has two neighbors other than itself.  As
shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
the number of particles in the speculative swarm.  The extra evaluations done
in speculative evaluation would instead be particles in regular PSO, so we
compare swarms of size $p_s$ in speculative evaluation with swarms of size
$8p_s$ in regular PSO.

We use a Random topology in our speculative algorithm, as the Complete topology
leads to an explosion of speculative evaluations.  However, if speculative
evaluation were not being performed, it is likely that the Complete topology
would be used, ignoring the communication cost of Complete over Random.  To be
fair in our comparisons, for the standard PSO algorithm we compare to both the
Random topology with two neighbors and to the Complete topology.  All functions
have 20 dimensions, and all results shown are averages over 20 runs.  We use
240 processors in our experiments, so speculative algorithms use 30 particles
and standard PSO uses 240 particles.  For ease of labeling, we call our
speculative algorithm SEPSO, the variant SEPSO Pick Best, and standard PSO just
PSO.

>>>>

\documentclass[letterpaper]{sig-alt-release2}

@@

@@

We also propose another way of making use of these speculative particles,
keeping the best position found instead of the position that PSO actually would
have taken.  We show that for a number of functions, speculative evaluation
gives dramatic improvements over adding additional particles to the swarm.

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05},
and algorithms based on Google's MapReduce framework \cite{mcnabb-cec07}.
Parallelizing the evaluation of the objective function can also be done in some
cases, though that is not an adaption of the PSO algorithm itself and thus is
not the focus of this paper.

@@

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait (e.g. a memory cache miss), guess which way the
branch will go and start executing.  If the processor guessed right, execution
is much farther ahead than if it had just idly waited on the branch.  If it
guessed wrong, execution restarts right where it would have been anyway.  This
is called speculative execution or branch prediction.

@@

Speculative evaluation is possible in PSO because the motion of particles
depends only on the \emph{positions} of the particle and its best neighbor, not
the \emph{value} of the objective function at those positions.  We do not have
to wait for a particle's function evaluation to complete in order to know where
it might move to next, so we can compute the set of all possible next positions
and send them to be evaluated along with the particle's current position.  Once
the current position of each particle has been evaluated, we can determine
which of the speculatively evaluated positions matches the branch taken by the
particle.  We have thus evaluated two iterations at once.  This can easily get
unwieldy if the set of next possible positions for each particle is large, but
a wise choice of topology limits that set to a reasonable size.

The original PSO can be reproduced exactly, two iterations at a time, using
this technique.  However, in strictly using speculative evaluation to reproduce
the original PSO, a lot of positions are speculatively evaluated whose values
get ignored after deciding which branch was the correct one.  A better use of
those speculative evaluations may be to incorporate the information into the
swarm, no longer exactly reproducing PSO but in some instances producing better
results.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
a multi-dimensional space by mimicking the swarming and flocking behavior of
birds and other animals. It is a social algorithm that depends on interaction
between particles to quickly and consistently find the optimal solution to a
given objective function.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
all possible combinations of updates that can be seen by a particle with two
neighbors other than itself, which is the case in all of the topologies we use.
If all of these speculative positions are evaluated in parallel along with the
original, non-speculative position of the particle at time $t-1$, then one of
the speculative positions has to be the correct one, and two iterations have
been evaluated at once.

\begin{table}
  \caption{All possible updates for a particle with two neighbors.}
  \label{tab:evals}
  \centering
  \begin{tabular}{cc}
	$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	\hline
	No update&No update\\
	\hline
	No update&Neighbor 1\\
	\hline
	No update&Neighbor 2\\
	\hline
	Updated&No update\\
	\hline
	Updated&Neighbor 1\\
	\hline
	Updated&Neighbor 2\\
	\hline
	Updated&The particle itself\\
	\hline
  \end{tabular}
\end{table}

Because $2(n+1)-1$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require
$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In that topology, 2 neighbors are chosen at random from the
swarm for each particle at each iteration.  Because a particle is also always
its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

@@

To aid in describing our methodology, we introduce a few terms.  A particle's
set of \emph{speculative children} is the set of all possible next iteration
states that a particle could have.  We use $\p_t$ to denote a particle at
iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
corresponding to one of the rows in Table~\ref{tab:evals}.  $\n_t$ is a
neighbor of particle $\p_t$.  Sets of particles are given by $\pset$, $\sset$,
or $\nset$, whereas single particles are simply $\p$, $\s$, or $\n$.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
a scheme, the master generates the positions to evaluate speculatively and
decides which one to accept for each particle.  The outline of the procedure is
given in \alg{centralized}.

@@

This is done by first updating each particle $\noeval{\p}_t$ to
$\nonbest{\p}_t$ and each speculative particle $\noeval{\s}_{t+1}$ to
$\nonbest{\s}_{t+1}$ by associating the value determined by the slave to the
particle it came from, and updating that particle's personal best if the value
reported is better than its previous personal best value.  Then each
$\nonbest{\p}_t$ must be updated to $\p_t$ by updating its neighborhood best
position and value with information about each of its neighbors.  This is
simply the original PSO algorithm, and corresponds to steps 1--5 in
\alg{centralized}.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depend entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected (step 7 in \alg{centralized}).

@@

The parent also needs to pass the value of the neighborhood best that the child
guessed.  The child only knows the position and needs the value in order to
make future comparisons between neighborhood best positions (step 8).

Upon picking the correct branch for each particle and updating the children's
personal best and neighborhood best value (from iteration $t$), the result is
the set $\nonbest{\pset}_{t+1}$, as the particles are now no longer
speculative.  What remains is to update the neighborhood best of those
particles from their neighbors (from iteration $t+1$), as above, to obtain
$\pset_{t+1}$.  That set of particles can subsequently be used to produce the
sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$ (steps 1 and 2 in
\alg{centralized}), and the process repeats itself.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those particles.
Messages such as values and positions for the neighborhood best are sent
between processors.  There may still be some machine that collects information
from all of the particles and outputs the result of the algorithm, though that
machine's importance is much less than in centralized algorithms.

To perform speculative evaluation in a distributed PSO algorithm, there must be
some way to have processors evaluate the speculative children of particles,
without giving the speculative particles the same treatment as actual
particles, as the speculative children only live for one iteration.  One way
that can be done is by assigning each particle a set of machines instead of a
single machine, and the particle directs its extra machines to evaluate its
speculative children.  However the distributed framework functions, the same
information needs to be passed between particles, and we describe here the
messages each particle needs to receive to perform speculative evaluation.

@@

In some distributed frameworks, like the MapReduce framework that we used in
our experiments, rounds of message passing can be expensive.  The method just
described uses three rounds of message passing for every two iterations
(corresponding to steps 2, 5 and 10 in \alg{centralized}).  Accordingly, we
describe a method that requires only one round of communication for each pair
of iterations, which happens at step 5 of \alg{centralized}.  Many more
messages are needed, but that is sometimes more desirable than synchronizing
all of the machines three times.

@@

With $\nonbest{\nset}_{t+1}$ and $\nonbest{\p}_{t+1}$, we can produce
$\p_{t+1}$.  This is shown in \fig{messages4}.  Note that we just combined
Figures~\ref{fig:messages2}~and~\ref{fig:messages3}, putting them together
to make $\p_{t+1}$, as all the particle needs is its neighborhood best to be
updated.

@@

This may seem like an inordinate amount of work, and with some distributed PSO
frameworks it is.  However, other parallel frameworks necessitate this type of
message passing, so we have described how speculative evaluation can be
performed in those circumstances.

@@

Performing speculative evaluation in PSO with a dynamic topology raises a few
issues of its own.  First, with neighbors changing each iteration, messages
that processors pass to their neighbors need to be sent to the correct
neighbors for each iteration.  In centralized algorithms and in distributed
algorithms with several rounds of message passing (the first method described
above), nothing needs to be changed.  However, in the case where processors
reconstruct information about neighboring particles from messages, those
messages need to be sent to the correct neighbors.  A particle cannot simply
send messages to its neighbors' neighbors' neighbors---it needs to send
messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
For every neighbor outward information is sent, the iteration also needs to be
incremented, as information about neighbors' neighbors is used during iteration
$t+1$, and information about neighbors' neighbors' neighbors is used to
reconstruct information about iteration $t+2$.  Also, this method of message
passing again requires the use of random seeds if the topology is random, so
that each processor computes the same neighbors for a particle as all other
processors.

@@

In a dynamic topology, a particle might not have information about the previous
positions of its neighbors at iteration $t$.  That means that its new
neighborhood best could come not only from its neighbors' positions at
iteration $t$, but also from their personal best from iteration $t-1$, as
neighbors' personal bests are what are used to update a particle's neighborhood
best.  That creates a problem for speculative evaluation---there are
potentially more than $2(n+1)-1$ possible next positions, increasing the amount
of work that must be done to perform the second iteration at the same time as
the first.

This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
available information about $\noeval{\nset}_{t+2}$ before producing
$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its 
neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
still only $2(n+1)-1$ possible next positions, and the problem is avoided.

@@

In performing speculative evaluation as we have described it, $2(n+1)-1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  It seems reasonable to try to make use of those
evaluations instead of throwing them away.  Making use of them changes the
behavior of PSO, instead of reproducing it exactly as the above method
explains, but the change could be an improvement.

@@

In our experiments we compared our speculative PSO algorithm to the standard
PSO algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  The speculative algorithm actually performs two iterations of
PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
evaluations.  For benchmark functions with very fast evaluations this may not
be the best use of parallelism in PSO.  But for many real-world applications,
the objective function takes on the order of minutes or more to evaluate; in
such cases our assumption is perfectly reasonable.

To use the same number of processors for each algorithm, speculative evaluation
needs a smaller swarm size than standard PSO.  For each of the topologies we
used, a particle has two neighbors other than itself.  As shown in
Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
the number of particles in the speculative swarm.  The extra evaluations done
in speculative evaluation would instead be particles in regular PSO, so we
compare swarms of size $p_s$ in speculative evaluation with swarms of size
$8p_s$ in regular PSO.

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion of speculative
evaluations.  However, if speculative evaluation were not being performed, it
is likely that the Complete topology would be used, ignoring the communication
cost of Complete over Random.  To be fair in our comparisons, for the standard
PSO algorithm we compare to both the Random topology with two neighbors and to
the Complete topology.  All functions have 20 dimensions, and all results shown
are averages over 20 runs.  We use 240 processors in our experiments, so
speculative algorithms use 30 particles and standard PSO uses 240 particles.
For ease of labeling, we call our speculative algorithm SEPSO, the variant
SEPSO Pick Best, and standard PSO just PSO.
